\chapter*{Acknowledgements}
\label{ch:acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

\begin{flushright}
{\footnotesize \textit {``Round up the usual suspects''}}
\end{flushright}

This thesis would not have been possible without the support, encouragement, and guidance of my supervisor, Chris Clarkson. I am grateful for his time and his patience with me as I stumbled my way through this PhD-- I know it hasn't always been easy. Thanks for not losing confidence in me, even when I lost my own. I would also like to express my deepest gratitude to Roy Maartens, for his guidance, sharing his expertise with me, and hosting me in Cape Town. I feel privileged having been able to spend time working in such an incredible environment.

Thanks to Alkistis Pourtsidou, for always listening and keeping me grounded, especially while writing this thesis; Phil Bull, for the many discussions about research, academia, and other; Richard Nelson, for encouragement and when I needed it; Tim Clifton, for his patience and advice; and all other lecturers and researchers in the Astronomy Unit, for making it such a great group to be part of. 

The work presented here would not have been possible without my fantastic collaborators, who taught me so much: Stefano Camera, Sheean Jolicoeur, and Obinna Umeh. I thank Z\'e Fonseca for advice on coding and life-- I promise I will be better at Python and less good at doubting myself in the future. 

My fellow PhD students past and present-- who have made being at Queen Mary so memorable-- thank you! Special thanks to John Ronayne for expertly taping bits of paper back together, Pedro Carrilho, for patiently answering all my ignorant questions about cosmology. My fellow cosmologists: Theo Anton, Louis Coates, Rebeca Carrillo, Jessie Durk, Pedro Fernandes, Jorge Fuentes, Kit Gallagher, Shailee Imrith, Fraser Kennedy, Charlie Pittordis, Paula Soares; and the non-cosmologists: Clark Baker, Ali Barlas, Kevin Chan, Alice Giroul, Paul Hallam, Francesco Lovascio, Sanson Poon, Jack Skinner, John Strachan, Domenico Trotta. Sanson, I apologise for turning the office into a jungle with a fly infestation-- despite my best efforts it is evident that I don't have green fingers. Callum, Rebeca, Francesco-- we made it! I know there is a bunch of you I am undoubtedly forgetting, thank you all. 

My friends near and far-- my Edinburgh friends, especially Ben Cullen, for always interesting discussions; Nicolas Angelides, for always being fabulous; and Nathan Johnson, for always being yourself. Francisco Vargas, thanks for teaching me the ins and outs of git over a pint, and all other computing advice ever. Thanks to David Grabner for listening, helping out, and hanging out; Lisa van Son, for always being there for me and coming to see me in every place I have lived; and Sergio Vargas for being supportive and performing Mathematica magic whenever I felt stuck.

Thanks to the GS gang (of the non-violent kind) for keeping me somewhat sane when the pandemic hit, and all the good times.

Finally, I am grateful for my family-- thanks to my wonderful parents and brother for supporting me, and welcoming me back with open arms when I tried to escape from plague island. Dankjulliewel! 
And Fran, thanks for your support, and for helping me get myself together, when I thought this thesis madness would never end. 



