\setlength{\tabcolsep}{0in}
\newcommand{\isep}{-2 pt}
\newcommand{\lsep}{-0.5cm}
\newcommand{\psep}{-0.6cm}
\renewcommand{\labelitemii}{$\circ$}
%


\college{Queen Mary, }
%
%
\department{Physics and Astronomy}
%
%
\supervisor{Dr. Chris Clarkson \& Dr. Alkistis Pourtsidou}
%
%
%
\title{Relativistic effects in the galaxy bispectrum}
% \title{On the relativistic bispectrum of large scale structure}
%\title{BS about the BS}
%
%
\author{Eline Maaike de Weerd}
\date{xxx 2021}% cth added
%Note Actual title format is done in icldt.cls
%
%
%\hypersetup{pdftitle={Non-linear effects in early Universe cosmology},pdfauthor={Pedro Carrilho}}
%

% Change as appropriate
\declaration{%
I, Eline Maaike de Weerd, confirm that the research included within this thesis is my own work or that where it has been carried out in collaboration with, or supported by others, that this is duly acknowledged below and my contribution indicated. Previously published material is also acknowledged below.
\newline
\newline
I attest that I have exercised reasonable care to ensure that the work is original, and does not to the best of my knowledge break any UK law, infringe any third party's copyright or other Intellectual Property Right, or contain any confidential material.
\newline
\newline
I accept that the College has the right to use plagiarism detection software to check the electronic version of the thesis.
\newline
\newline
I confirm that this thesis has not been previously submitted for the award of a degree by this or any other university. The copyright of this thesis rests with the author and no quotation from it or information derived from it may be published without the prior written consent of the author.
\newline
\newline
Details of collaboration and publications:
Part of this work has been done in collaboration with Stefano Camera, Chris Clarkson, Sheean Jolicoeur, Roy Maartens, and Obinna Umeh. It is based on the following publications, all of which I have contributed to:
\begin{itemize}
	\item \underline{The dipole of the galaxy bispectrum} \\
	C. Clarkson, E.M. De Weerd, S. Jolicoeur, R. Maartens, O. Umeh \\
	Published: \textit{MNRAS Letters 486 (2019) L101},	arXiv: 1812.09512 [astro-ph.CO]
	\item \underline{Detecting the relativistic galaxy bispectrum} \\
	R. Maartens, S. Jolicoeur, O. Umeh, E.M. De Weerd, C. Clarkson, S. Camera\\
	Published: \textit{JCAP03(2020)065},	arXiv: 1911.02398 [astro-ph.CO]
	\item \underline{Multipoles of the relativistic galaxy bispectrum} \\
	E.M. De Weerd, C. Clarkson, S. Jolicoeur, R. Maartens, O. Umeh \\
	Published: \textit{JCAP05(2020)018},	arXiv: 1912.11016 [astro-ph.CO]
	\item \underline{Detecting the relativistic bispectrum in 21cm intensity maps} \\
	S. Jolicoeur, R. Maartens, E.M. De Weerd, O. Umeh, C. Clarkson, S. Camera \\
	Published: \textit{JCAP06(2021)039},	arXiv: 2009.06197 [astro-ph.CO]
	\item \underline{Local primordial non-Gaussianity in the relativistic galaxy bispectrum} \\
	R. Maartens, S. Jolicoeur, O. Umeh, E.M. De Weerd, C. Clarkson \\
	Published: \textit{JCAP04(2021)013},	arXiv: 2011.13660 [astro-ph.CO]
\end{itemize}
\vfill
\noindent Signature: Eline Maaike de Weerd
\newline
Date: xx.xx.2021}
\thispagestyle{empty}


\maketitle
